# Formula 1 winner prediction.

### DATA CLEANING AND PROCESSING

```{r}
head(results)
```

```{r}
head(status)
```

```{r}
head(drivers)
```

```{r}
head(races)
```

```{r}
head(constructors)
```

```{r}
head(races)
```

```{r}
head(constructors)
```

```{r}
head(circuits)
```

```{r}
head(constructor_standings)
```

Dropping uwnanted columns from results.

```{r}
results = results[-c(12:15)]
colnames(results)[7] = 'finalposition'
```

Merging status in results

```{r}
df.results.merged = merge(results, status, by = c('statusId'))
df.results.merged = df.results.merged %>% relocate(statusId, .after = last_col())
```

Merging drivers in df results

```{r}
#pasting forename and surname in a new column 'fullname' in drivers df.
drivers$fullname = paste(drivers$forename, drivers$surname)
names(drivers$nationality) = 'driv_nationality' #renaming driver's nationality column
drivers_copy = drivers_copy[-c(3, 4, 9)]  #creating with only columns to be merged with results


#merging
df.results.merged = merge(df.results.merged, drivers_copy, by = c('driverId'))

```

Merging races with df.results.merged

```{r}
races_copy = races[, c(1,3:6)]
df.results.merged = merge(df.results.merged, races_copy, by = c('raceId'))

```

Merging constructors with df results

```{r}
constructors_copy = constructors[, c(1,3)]
colnames(df.results.merged)[22] = 'const_name'

df.results.merged = merge(df.results.merged, constructors_copy, by = 'constructorId')
df.results.merged = df.results.merged[,-c(20)]

```

Merging circuits with df results.

```{r}
circuits_copy = circuits[, c(1,3)]

df.results.merged = merge(df.results.merged, circuits_copy, by = c('circuitId'))
```

Merging constructor standings with df.result for importing constructor wins and points

```{r}
const_standing_copy = constructor_standings[, c(2,3,4,7)]
colnames(const_standing_copy)[c(3:4)] = c('const_points', 'const_wins')

df.results.merged = merge(df.results.merged, const_standing_copy, by = c('raceId', 'constructorId'))
```

PROCESSING some data columns.

```{r}
#casting fastestLapSpeed to numeric
df.results.merged$fastestLapSpeed = as.numeric(df.results.merged$fastestLapSpeed)

#substitute NA's in fastestSpeed with 0.
df.results.merged$fastestLapSpeed[is.na(df.results.merged$fastestLapSpeed)] <- 0

#Calculating drivers age and adding a relative column to the dataset
df.results.merged$driver_age = as.integer(format(df.results.merged$date, format="%Y"))- as.integer(format(df.results.merged$dob, format="%Y"))

#convert fastest lap time in number of milliseconds 
df.results.merged$fastestLap_ms = as.numeric(lubridate::ms(as.character(df.results.merged$fastestLapTime)))*1000
df.results.merged$fastestLap_ms[is.na(df.results.merged$fastestLap_ms)] <- 0
df.results.merged = df.results.merged[, -c(11)]  #dropping column of old fastestLapTime 


```

### EDA and plots.

```{r}

hist(merged_numeric$driver_age, main = paste("Histogram of driver age"))

```

```{r}
boxplot(df.results.merged[22], horizontal = TRUE, notch = TRUE, xlab = "DRIVERS' AGE")
```



```{r}
par(mar=c(3, 7, 0, 0))
barplot(
  table(df.results.merged$name),
  horiz = TRUE, 
  las=1, 
  cex.names=0.4
  )
```


Number of collisions for each circuit.

```{r}
#barplot for collisions
par(mar=c(3, 7, 2, 1))

#barplot showing number of collisions for each circuit
barplot(
  table(df.results.merged[df.results.merged$status == 'Collision', 'name']), 
  horiz = TRUE, 
  las=1, 
  cex.names=0.4
  )
```

Accident ratio for each circuit.

```{r}
#get circuit names for which there was a collision
accident_names = names(table(df.results.merged[df.results.merged$status == 'Accident', 'name']))

#barplot showing accident ratio (n^Accident/n^races per circuit)
par(mar=c(3, 7, 2, 1))
barplot(
  (table(df.results.merged[df.results.merged$status == 'Accident', 'name']))/(table(df.results.merged[, 'name'])[accident_names]),
  horiz = TRUE, 
  las=1, 
  cex.names=0.4
  )

```

Recall function for winner age for.

```{r}
winner_Age()
df_year_winner = winner_Age()

```

```{r}
df_year_winner
```

```{r}
#plot for winner's age over time
ggplot(df_year_winner, aes(x = X1, y = X5)) + 
  labs(x = 'Years', y = 'Age') +
  ggtitle('Winning driver age for every year') +
  geom_bar(stat = "identity") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  geom_abline(slope = 0, intercept = mean(df_year_winner$X5), color = 'red', size = 1.5) + theme_bw() + 
  geom_text(aes(label = X5), vjust = -1, size=5)
```

CORRELATION
```{r}
#CORRELATION MATRIX for numerical features
corr_matrix = cor(df.results.merged[c(7,8,9,10,11,16,20,21,22,23)])
corrplot(corr_matrix)
```



### DEFINING AND TRAINING ML MODELS FOR PREDICTION. 









